Unsupervised embedding of trajectories captures the latent structure of mobility

### 2020-10-05 paper/current/main.tex
edits to just before conlusion

Prestige hierarchy is known to underpin the dynamics of scientific mobility, where researchers tend to move to similar or less prestigious organizations~\autocite{deville2014career, clauset2015hierarchy}. Could the embedding space, to which no explicit prestige information is given, encode a prestige hierarchy?
We test this by exploiting the geometric properties of the embedding space with SemAxis~\autocite{an2018semaxis}.
Here, we use SemAxis to operationalize the abstract notion of academic prestige, defining an axis in the embedding space where poles are defined using known high- and low-ranked universities.
As an external measure of prestige, we use the Times Ranking of World Universities (we also use research impact from the Leiden Rankings~\autocite{waltman2012leidenrankings}, see \SI);
the high-rank pole is defined as the average vector of the top five U.S. universities according to the rankings, whereas the low-ranked pole is defined using the five bottom-ranked (geographically-matched by U.S. census region) universities.
We derive an embedding-based ranking for universities based on the spectrum from the high-ranked to low-ranked poles (see Data and Methods).

The embedding space encodes the prestige hierarchy of U.S. universities that are coherent with real-world university rankings.
The embedding-based ranking is strongly correlated with the Times ranking (Spearman's $\rho = 0.73$, Fig.~\ref{fig:semaxis}a).
For reference, the correlation between the Times ranking and the publication impact scores from the Leiden Ranking~\autocite{waltman2012leidenrankings}, a bibliometrically-based university ranking is 0.87 (Spearman's $\rho$, Fig.~\ref{fig:semaxis}b).
The correlation between the embedding-based ranking and the Times ranking is robust regardless of the number of organizations used to define the axes (Fig. ~\ref{fig:supp:semaxis_compare}), such that even using only the single top-ranked and bottom-ranked universities produces a ranking that is significantly correlated with the Times ranking (Spearman's $\rho = 0.46$, Fig.~\ref{fig:supp:semaxis_compare}a).
The correlation is also comparable to network-driven measures such as degree strength (sum of edge weights, Spearman's $\rho = 0.73$) and eigenvector centrality (Spearman's $\rho = 0.76$, see \SI). 
The strongest outliers that were ranked more highly in the Times ranking than in the embedding-based ranking tend to be large state universities such as Arizona State University and the University of Florida.
Those ranked higher in the embedding-based ranking tend to be relatively-small universities near major urban areas such as the University of San Francisco and the University of Maryland Baltimore County, possibly reflecting exchanges of scholars with nearby high-ranked institutions at these locations.
In sum, our result suggests that the embedding space is capable of capturing information about academic prestige, even when the representation is learned using data without explicit information on the direction of mobility (as in other formal models~\autocite{clauset2015hierarchy}), or prestige.

The axes can further be visualized to examine the relative position of organizations along the prestige axis, and along a geographic axis between California and Massachusetts.
Prestigious universities such as Columbia, Stanford, MIT, Harvard, and Rockefeller are positioned towards the top of the axis (Fig.~\ref{fig:semaxis}c).
Universities at the bottom of this axis tend to be regional universities with lower national profiles (yet still ranked by Times Higher Education) and with more emphasis on teaching, such as Barry University and California State University at Long Beach.
By projecting other types of organizations onto the prestige axis, SemAxis offers a new means of reasoning about the prestige of organizations for which rankings are often low-resolution, incomplete, or entirely absent, such as regional and liberal arts universities, seminaries, musical schools, community colleges (Fig.~\ref{fig:semaxis}d), government organizations (Fig.~\ref{fig:semaxis}d), and research institutes (Fig.~\ref{fig:semaxis}e).
Their estimated prestige has a significant correlation with their citation impact (Fig.~\ref{fig:supp:sector_semaxis_impact}).

@@

We also find that the size (L2 norm) of the organization embedding vectors provides insights into the characteristics of organizations.
Up to a point (around 1,000 researchers), the size of U.S. organization's vectors tends to increase proportionally to the number of researchers (both mobile and non-mobile) with published work;
these organizations are primarily teaching-focused institutions, agencies, and hospitals that either are not ranked or have a low ranking.
However, at around 1,000 researchers, the size of the vector decreases as the number of researchers increases.
These organizations are primarily research-intensive and prestigious universities with higher rankings, research outputs, R\&D funding, and doctoral students (Fig.~\ref{fig:supp:length_vs_metaInfo}).
A similar pattern has been observed in applications of neural embedding to natural language, in which the size of word vectors was found to represent the word's \emph{specificity}.
Roughly speaking, for word embeddings, the size of a vector indicates a specificity, i.e., the word associated with the vector frequently co-appears with particular context words\autocite{schakel2015measuring}. 
If the word in question is universal, appearing frequently in many different contexts, it would not have a large norm due to a lack of strong association with a particular context.
Likewise, an organization such as Harvard with a small norm appears in many contexts, alongside many different organizations in affiliation trajectories---it is well-connected.
The concavity emerges in part from the relationship between the size of the vector and the expected connectedness of the organization given its size (Fig.~\ref{fig:length}a, inset, $R^{2} = 0.17$).
Large, prestigious, and well-funded research universities such as Princeton and Harvard have smaller vector norms because they appear in many different contexts compared to more teaching-focused organizations such as NY Medical College, and the University of Michigan at Flint.
Some universities, such as the University of Alaska at Fairbanks, have considerably small vectors, which may be a result of its remote location and unique circumstances.

We report that this curve is almost universal across many countries. 
For instance, China's curve closely mirrors that of the United States (Fig.~\ref{fig:length}b).
Smaller but scientifically advanced countries such as Australia, and other populous countries such as Brazil also exhibit curves similar to the United States (Fig.~\ref{fig:length}b, inset), though their exact position and size vary.
Other nations exhibit different curves, such as Egypt which has an``S''-shaped curve, meaning that their largest organizations are not as well connected as other countries' largest organizations. 
Similar patterns can be found across many of the 30 of the countries with the most total researchers (Fig.~\ref{fig:concave30}), however, we can only speculate on what their variations mean (see \SI for more discussion).

%
% Figure - Boomerang
%
\begin{figure}[h!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/Math/boomerang.pdf}
	\caption{
		\textbf{Size of organization embedding vectors captures prestige and size of organizations.}
		\textbf{a.} Size (L2 norm) of organization embedding vectors compared to the number of researchers for U.S. universities.
		Color indicates the rank of the university from the Times ranking, with 1 being the highest ranked university.
		Uncolored points are universities not listed on the Times ranking.
		A convex-shape emerges, wherein larger universities tend to be more distant from the origin (large L2 norm); however, the more prestigious universities tend to have smaller L2 norms.
		The L2 norm is correlated weakly with the ratio of organizations' number of connections (non-zero flows) to the number of researchers ($R^{2} = 0.17$, \textbf{inset}).
		The red line is the line of the best fit, and 99\% confidence intervals are shown, though are too small to be visible.
		\textbf{b.} We find a similar convex-curve pattern across many countries such as the United States, China, Australia, Brazil, and others (\textit{inset}, and Fig.~\ref{fig:concave30}). 
		Some countries exhibit variants of this pattern, such as Egypt, which is missing the right side of the curve.
		The loess regression lines are shown for each selected country, and for the aggregate of remaining countries, with 99\% confidence intervals.
		Loess lines are also shown for organizations in Australia, Brazil, and Egypt (\textbf{inset}).
	}
	\label{fig:length}
\end{figure}

>>>>

Prestige hierarchy is known to underpin the dynamics of scientific mobility, where researchers tend to move to similar or less prestigious organizations~\autocite{deville2014career, clauset2015hierarchy}. Could the embedding space, to which no explicit prestige information is given, encode a prestige hierarchy?
This question is tested by exploiting the geometric properties of the embedding space with SemAxis~\autocite{an2018semaxis}.
Here, we use SemAxis to operationalize the abstract notion of academic prestige, defining an axis in the embedding space where poles are defined using known high- and low-ranked universities.
As an external measure of prestige, we use the Times Ranking of World Universities (we also use research impact from the Leiden Rankings~\autocite{waltman2012leidenrankings}, see \SI);
the high-rank pole is defined as the average vector of the top five U.S. universities according to the rankings, whereas the low-rank pole is defined using the five bottom-ranked (geographically-matched by U.S. census region) universities.
We derive an embedding-based ranking for universities based on the geometrical spectrum from the high-ranked to low-ranked poles (see Data and Methods).

The embedding space encodes the prestige hierarchy of U.S. universities that are coherent with real-world university rankings.
The embedding-based ranking is strongly correlated with the Times ranking (Spearman's $\rho = 0.73$, Fig.~\ref{fig:semaxis}a).
For reference, the correlation between the Times ranking and the publication impact scores from the Leiden Ranking~\autocite{waltman2012leidenrankings}, a bibliometrically-based university ranking is 0.87 (Spearman's $\rho$, Fig.~\ref{fig:semaxis}b).
The correlation between the embedding-based ranking and the Times ranking is robust regardless of the number of organizations used to define the axes (Fig. ~\ref{fig:supp:semaxis_compare}), such that even using only the single top-ranked and bottom-ranked universities produces a ranking that is significantly correlated with the Times ranking (Spearman's $\rho = 0.46$, Fig.~\ref{fig:supp:semaxis_compare}a).
The correlation is also comparable to more direct measures such as node strength (sum of edge weights, Spearman's $\rho = 0.73$) and eigenvector centrality (Spearman's $\rho = 0.76$, see \SI) from the mobility network. 
The strongest outliers that were ranked more highly in the Times ranking than in the embedding-based ranking tend to be large state universities such as Arizona State University and the University of Florida.
Those ranked higher in the embedding-based ranking tend to be relatively-small universities near major urban areas such as the University of San Francisco and the University of Maryland Baltimore County, possibly reflecting exchanges of scholars with nearby high-ranked institutions at these locations.
In sum, our result suggests that the embedding space is capable of capturing information about academic prestige, even when the representation is learned using data without explicit information on the direction of mobility (as in other formal models~\autocite{clauset2015hierarchy}), or prestige.

The axes can be visualized to examine the relative position of organizations along the prestige axis, and along a geographic axis between California and Massachusetts.
Prestigious universities such as Columbia, Stanford, MIT, Harvard, and Rockefeller are positioned towards the top of the axis (Fig.~\ref{fig:semaxis}c).
Universities at the bottom of this axis tend to be regional universities with lower national profiles (yet still ranked by Times Higher Education) and with more emphasis on teaching, such as Barry University and California State University at Long Beach.
By projecting other types of organizations onto the prestige axis, SemAxis offers a new means of reason about the prestige of organizations for which rankings are often low-resolution, incomplete, or entirely absent, such as regional and liberal arts universities, seminaries, musical schools, community colleges (Fig.~\ref{fig:semaxis}d), government organizations (Fig.~\ref{fig:semaxis}d), and research institutes (Fig.~\ref{fig:semaxis}e).
Their estimated prestige is speculative, though we find that it significantly correlates with their citation impact (Fig.~\ref{fig:supp:sector_semaxis_impact}).

@@

We also find that the size (L2 norm) of the organization embedding vectors provides insights into the characteristics of organizations.
Up to a point (around 1,000 researchers), the size of U.S. organization's vectors tends to increase proportionally to the number of researchers (both mobile and non-mobile) with published work;
these organizations are primarily teaching-focused institutions, agencies, and hospitals that either are not ranked or have a low ranking.
However, at around 1,000 researchers, the size of the vector \emph{decreases} as the number of researchers increases.
These organizations are primarily research-intensive and prestigious universities with higher rankings, research outputs, R\&D funding, and doctoral students (Fig.~\ref{fig:supp:length_vs_metaInfo}).
A similar pattern has been observed in applications of neural embedding to natural language, in which the size of word vectors was found to represent the word's \emph{specificity}.
Roughly speaking, for word embeddings, the size of a vector reflects its specificity, i.e., the word associated with the vector frequently co-appears with particular context words\autocite{schakel2015measuring}. 
If the word in question is universal, appearing frequently in many different contexts, it would not have a large norm due to a lack of strong association with a particular context.
Likewise, an organization such as Harvard with a small norm appears in many contexts, alongside many different organizations in affiliation trajectories---it is well-connected.
The concavity emerges in part from the relationship between the size of the vector and the expected connectedness of the organization given its size ($R^{2} = 0.17$).
Large, prestigious, and well-funded research universities such as Princeton and Harvard have smaller vector norms because they appear in many different contexts compared to more teaching-focused organizations such as NY Medical College, and the University of Michigan at Flint.
Some universities, such as the University of Alaska at Fairbanks, have considerably small vectors, which may be a result of its remote location and unique circumstances.

We report that this curve is almost universal across many countries. 
For instance, China's curve closely mirrors that of the United States (Fig.~\ref{fig:length}b).
Smaller but scientifically advanced countries such as Australia, and other populous countries such as Brazil also exhibit curves similar to the United States (Fig.~\ref{fig:length}b, inset), though their exact position and size vary.
Other nations exhibit different curves, such as Egypt which has an``S''-shaped curve, meaning that their largest organizations are not as well connected as other countries' largest organizations. 
Similar patterns can be found across many of the 30 of the countries with the most total researchers (Fig.~\ref{fig:concave30}) see \SI{} for more discussion).

%
% Figure - Boomerang
%
\begin{figure}[h!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/Math/boomerang.pdf}
	\caption{
		\textbf{Size of organization embedding vectors captures prestige and size of organizations.}
		\textbf{a.} Size (L2 norm) of organization embedding vectors compared to the number of researchers for U.S. universities.
		Color indicates the rank of the university from the Times ranking, with 1 being the highest ranked university.
		Uncolored points are universities not listed on the Times ranking.
		A convex-shape emerges, wherein larger universities tend to be more distant from the origin (large L2 norm); however, the more prestigious universities tend to have smaller L2 norms.
		\textbf{b.} We find a similar convex-curve pattern across many countries such as the United States, China, Australia, Brazil, and others (\textit{inset}, and Fig.~\ref{fig:concave30}). 
		Some countries exhibit variants of this pattern, such as Egypt, which is missing the right side of the curve.
		The loess regression lines are shown for each selected country, and for the aggregate of remaining countries, with 99\% confidence intervals.
		Loess lines are also shown for organizations in Australia, Brazil, and Egypt (\textbf{inset}).
	}
	\label{fig:length}
\end{figure}
